---
title: "Assignment 2: Coding Basics"
author: "Yixin Fang"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. 
sequence <- seq(1,100,4) #Create the sequence and name it 
sequence #Show the result of the sequence 

#2. 
mean_sequence <- mean(sequence) #Compute the mean of the sequence
median_sequence <- median(sequence) #Compute the median of the sequence
mean_sequence #Show the mean
median_sequence #Show the median

#3. 
mean_sequence > median_sequence #See if the mean is larger than the median. The answer is FALSE

```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
#5.
names <- c('Hal','Moss','Sam','Shaw') #Character
scores <- c(98, 80, 76, 43) #Numeric
results <- c(TRUE, TRUE, TRUE, FALSE) #Logical

#7. 
student_test_results <- data.frame(names,scores,results)

#8.
colnames(students_test_results) <- c('student_names', 'test_scores', 'test_results')

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: Data frame can have different types of data while the data in matrix are usually the same type.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}
passfail <- ifelse(scores < 50, FALSE, TRUE)
print(passfail)

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer:'ifelse' worked. Because the input is a vector, and 'ifelse' is the vectorization of the 'if...else' function. 
